### 2022-08-20 main/chapter_searches.tex
autocheck; shrink xkcd to fit page

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside of their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but calculation with their moments derives
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
exactly if independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, the
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

@@

Logic is an important application of probability for which $a$ and $b$ are
interpreted as logical propositions, and
$\prob{b}{a}$ is the degree to which $a$ implies $b$, with the limiting cases
that $\prob{b}{a} = 1$ when $a$ implies $b$ and $\prob{b}{a} = 0$ when $a$
disproves $b$.
Indeed Cox's theorem derives the laws of probability by requiring compatibility
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic and it works \emph{for tea, too}.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.55\textwidth]{figures/searches_significant_shrink.png}
\\
\begin{footnotesize}
`So, uh, we did the green study again and got no link. It was probably a{-}{-}'
\\
`RESEARCH CONFLICTED ON GREEN JELLY BEAN ACNE LINK; MORE~STUDY~RECOMMENDED!'
\end{footnotesize}
\caption[
``Significant'' by Randall~Munroe
]{%
``Significant'' by Randall~Munroe~\cite{xkcd2011significant}.
}
\label{fig:searches_significant}
\end{figure}

@@

To model the consequences of these many factors of uncertainty, each factor is
assigned a real-valued parameter which when varied changed the expected yields
in the hisogram bins of the search.
In this way, we form a statistical model which is a function
\begin{equation}
H_{\!f}(\vec \theta) \rightarrow (\vec \mu, \log f)
\end{equation}
from the parameters $\theta$ to Poisson expectations $\vec\mu$ the histogram
bins and a regularization term $\log f$ which is maximized in fitting.

>>>>

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but calculation with their moments derives
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
exactly if independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, the
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

@@

Logic is an important application of probability for which $a$ and $b$ are
interpreted as logical propositions, and
$\prob{b}{a}$ is the degree to which $a$ implies $b$, with the limiting cases
that $\prob{b}{a} = 1$ when $a$ implies $b$ and $\prob{b}{a} = 0$ when $a$
disproves $b$.
Indeed, Cox's theorem derives the laws of probability by requiring compatibility
with Boolean logic~\cite{
cox1946probability,
cox1961algebra,
garrett1998nand,
jaynes2003probability,
keynes1920treatise
}.
Probability works for logic, and it works \emph{for tea, too}.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.46\textwidth]{figures/searches_significant_shrink.png}
\\
\begin{footnotesize}
`So, uh, we did the green study again and got no link. It was probably a{-}{-}'
\\
`RESEARCH CONFLICTED ON GREEN JELLY BEAN ACNE LINK; MORE~STUDY~RECOMMENDED!'
\end{footnotesize}
\caption[
``Significant'' by Randall~Munroe
]{%
``Significant'' by Randall~Munroe~\cite{xkcd2011significant}.
}
\label{fig:searches_significant}
\end{figure}

@@

\TODO{minimize energy not max logf}
To model the consequences of these many factors of uncertainty, each factor is
assigned a real-valued parameter, which when varied changed the expected yields
in the histogram bins of the search.
In this way, we form a statistical model which is a function
\begin{equation}
H_{\!f}(\vec \theta) \rightarrow (\vec \mu, \log f)
\end{equation}
from the parameters $\theta$ to Poisson expectations $\vec\mu$ of the histogram
bins and a regularization term $\log f$ that is maximized in fitting.
